---
title: "PopHIVE_Style_Guide"
output: html_document
date: "2026-02-08"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This style guide was created to ensure a consistent visual identity
across PopHIVE products. Please see the instructions below to apply our
pre-written functions to your data visualizations.

## Setup

### Load required packages

```{r echo=FALSE message=FALSE}
#load required packages
library(tidyverse)
library(ggplot2)
library(sf)
library(showtext)
library(scales)
library(maps)
library(ggstar)
```

### Load all functions

All necessary functions are located in PopHIVE Insights repository:
PopHIVE -\> Insights -\> PopHIVE_style_guide. Load these functions at
the start of your session with the following code:

```{r echo=FALSE message=FALSE}
source("pophive_style_functions.R")
```

There are functions for 5 different types of graphs included:\
\* choropleth map
\* bubble map\
\* line graph\
\* dot plot\
\* bar/ column chart

## Examples

### Choropleth Map

For the function to work correctly, your data must include a shapefile,
which you can load with the "sf" package. Follow this format (the
italicized text is what you will need to change):

ggplot(*data*) +
  geom_sf(aes(fill = *value_variable*), color = "gray", size = 0.3) +
  scale_fill_pophive_map(legend_title = "*Legend Title*") +
  theme_pophive_map() +
    labs(
      title = "*Title*",
      subtitle = "*Subtitle*",
      caption = "*Caption*")  

```{r}
# Example 

states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
states$value <- runif(nrow(states), 0, 100)

ggplot(states) +
  geom_sf(aes(fill = value), color = "gray", size = 0.3) +
  scale_fill_pophive_map(legend_title = "Population Density") +
  theme_pophive_map() +
  labs(
    title = "US Population Density by State",
    subtitle = "Example choropleth map with custom PopHive theme",
    caption = "Data Source: sf package"
  )
```

### Bubble Map


### Line Graph 

Note that for line graphs, you will need special helper functions for the color and linetype aesthetics. 

ggplot(*date*, aes(x = *time_variable*, y = *value*, color = *category*, linetype = *category*)) +
  geom_line(linewidth = 1) +
  scale_x_date(date_breaks = "*frequency*", date_labels = *"%b %d, %Y"*) + 
  scale_y_continuous(labels = scales::*scale_type*) +
  scale_color_pophive_line() +
  scale_linetype_pophive_line() +
  theme_pophive_line() +
  labs(
    title = "*Title*",
    subtitle = "*Subtitle*",
    x = "*X Axis Title*",
    y = "*Y Axis Title*"
  )

```{r}
df_line <- data.frame(
  year = rep(2015:2024, 3),
  value = c(runif(10, 20, 50), runif(10, 30, 60), runif(10, 25, 55), runif(10, 20, 50), runif(10, 30, 60), runif(10, 25, 55)),
  category = rep(c("Category A", "Category B", "Category C", "Category D", "Category E", "Category F"), each = 10)
)

ggplot(df_line, aes(x = year, y = value, color = category, linetype = category)) +
  geom_line(linewidth = 0.7) +
  scale_color_pophive_line() +
  scale_linetype_pophive_line() +
  theme_pophive_line() +
  labs(
    title = "Trends Over Time",
    subtitle = "Example line graph with custom PopHive theme",
    caption = "Data source: sample data",
    x = "Year",
    y = "Value"
  )
```


### Dot Plot

ggplot(*data*, aes(x = *value*, y = *category*)) +
  ggstar::geom_star(starshape = "hexagon", color = "#4e98e4", size = 4, fill = "#4e98e4") +
  theme_pophive_dot() +
  labs(
    title = "*Title*",
    subtitle = "*Subtitle*",
    caption="*Caption*",
    x = "*X Axis Title*",
    y = "*Y Axis Title*"
  )

```{r}
df_dot <- data.frame(
  category = c("Category A", "Category B", "Category C", "Category D", "Category E"),
  value = c(45, 67, 23, 89, 56)
)

ggplot(df_dot, aes(x = value, y = reorder(category, value))) +
  ggstar::geom_star(starshape = "hexagon", color = "#4e98e4", size = 4, fill = "#4e98e4") +
  theme_pophive_dot() +
  labs(
    title = "Category Comparison",
    subtitle = "Example horizontal dot plot with custom PopHive theme",
    caption="Data source: sample data",
    x = "Value",
    y = "Category"
  )
```


### Bar/ Column Chart

ggplot(*df_col*, aes(x = *category*, y = *value*, fill = *group*)) +
  geom_col(position = "*dodge/stack/fill*", width = 0.7) +
  scale_fill_pophive_col() +
  theme_pophive_col() +
  labs(
    title = "*Title*",
    subtitle = "*Subtitle*",
    caption = "*Caption*",
    x = "*X Axis Title*",
    y = "*Y Axis Title*"
  )

```{r}
df_col <- data.frame(
  category = rep(c("Q1", "Q2", "Q3", "Q4"), 3),
  value = c(runif(4, 20, 50), runif(4, 30, 60), runif(4, 25, 55)),
  group = rep(c("Product A", "Product B", "Product C"), each = 4)
)

ggplot(df_col, aes(x = category, y = value, fill = group)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_pophive_col() +
  theme_pophive_col() +
  labs(
    title = "Quarterly Sales by Product",
    subtitle = "Sales performance across different products and quarters",
    caption = "Data source: sample data",
    x = "Quarter",
    y = "Sales (thousands)"
  )
```

